---
title: "Logical Operators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{logicals-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

To start, load the package.

```{r setup}
library(extraoperators)
```

# Logical Comparisons

This section covers basic logical comparisons and shows how they might
be done in base `R` versus using the `extraoperators` package. Many of
these are quite simple, but are defined so that later operators are
possible.

First let's define our "data" as some numbers stored in `sample_numbers`.

```{r}

sample_numbers <- c(9, 1, 5, 3, 4, 10, 99)

``` 

Now we can do a series of simple logical comparisons which return a
logical vector of `TRUE` or `FALSE`.

```{r}

## base R: greater than 3?
sample_numbers > 3

## base R: greater than or equal to 3?
sample_numbers >= 3

## base R: less than 3?
sample_numbers < 3

## base R: less than or equal to 3?
sample_numbers <= 3

```

Unfortunately, we cannot use `<` or `>` in custom operators, so we use
the substitutions: `g = >` and `l = <` and `e = =`. So that `ge = <=`
etc.

```{r}

## extraoperators: greater than 3?
sample_numbers %g% 3

## extraoperators: greater than or equal to 3?
sample_numbers %ge% 3

## extraoperators: less than 3?
sample_numbers %l% 3

## extraoperators: less than or equal to 3?
sample_numbers %le% 3

```

So far there is no real gain in using `extraoperators` but this changes
for more complex operations. What if we want to know if our values
fall within some range? This is a fairly common task, such as saying
that valid ages must be between 0 and 100 years.

```{r} 

## base R: greater than 3 and less than 10?
sample_numbers > 3 & sample_numbers < 10

## base R: greater than or equal to 3 and less than 10?
sample_numbers >= 3 & sample_numbers < 10

## base R: greater than 3 and less than or equal to 10?
sample_numbers > 3 & sample_numbers <= 10

## base R: greater than or equal to 3 and less than or equal to 10?
sample_numbers >= 3 & sample_numbers <= 10

```

Base `R` accomplishes this through chaining of
operations. `extraoperators` has built in range operators.


```{r} 

## extraoperators: greater than 3 and less than 10?
sample_numbers %gl% c(3, 10)

## extraoperators: greater than or equal to 3 and less than 10?
sample_numbers %gel% c(3, 10)

## extraoperators: greater than 3 and less than or equal to 10?
sample_numbers %gle% c(3, 10)

## extraoperators: greater than or equal to 3 and less than or equal to 10?
sample_numbers %gele% c(3, 10)

```

Finally, `extraoperators` includes a `not in` operator, `%!in%`.

```{r}

## base R: not in 3 or 10
!sample_numbers %in% c(3, 10)

## extraoperators: not in 3 or 10
sample_numbers %!in% c(3, 10)

``` 

The next sections show a few examples of these operators augmented by
prefixes: `?`, `s` and `a`.

# Indices (Which Values?)

Sometimes we want to use a logical comparison and identify indices,
such as to use in a loop. `extraoperators` does this by prefixing
operators with `?` for "which".

```{r}

## base R: what are the indices that match 3 and 10?
which(sample_numbers %in% c(3, 10))

## extraoperators: what are the indices that match 3 and 10?
sample_numbers %?in% c(3, 10)

## base R: what are the indices for numbers between 3 and 10?
which(sample_numbers > 3 & sample_numbers < 10)

## extraoperators: what are the indices for numbers between 3 and 10?
sample_numbers %?gl% c(3, 10)

``` 

This can be readily incorporated in other code for further processing.

# Subsetting

Another fairly common task is selecting only certain observations.
For example, we might want to calculate the average of numbers within
a plausible range (e.g., excluding outliers).
In `extraoperators` subsetting is done by adding an `s` prefix.

```{r}

## base R: subset to only numbers between 3 and 10
mean(subset(sample_numbers, sample_numbers > 3 & sample_numbers < 10))

## or equivalently 
mean(sample_numbers[sample_numbers > 3 & sample_numbers < 10])

## extraoperators: subset to only numbers between 3 and 10
mean(sample_numbers %sgl% c(3, 10))

```

Subsetting can be especially useful in quick exploratory
analyses. Graphs are easily hard to read if there are extreme
values. Subsetting makes it fast to "zoom in" on a specific range.

# All (or None)

Finally, you might have some quality controls in place for data. For
example asserting that all ages are between 0 and 100. In `extraoperators`
this is done by adding the prefix `a`.

```{r}

## base R: are all numbers between 0 and 10?
all(sample_numbers > 0 & sample_numbers < 10)

## extraoperators: are all numbers between 0 and 10?
sample_numbers %agl% c(0, 10)

## extraoperators: are all numbers between 0 and 100?
sample_numbers %agl% c(0, 100)

```

If you want to know the opposite, are no numbers between 0 and 100, we
can negate the whole operation.

```{r}

## extraoperators: are NO numbers between 0 and 100?
!sample_numbers %agl% c(0, 100)

## extraoperators: are NO numbers between 55 and 60?
!sample_numbers %agl% c(55, 60)

```
